Stochastic Blockmodels for Relational Event Data

### 2012-03-30 docs/submission.tex
Update split merge description.

>>>>

IDEA: References are Dahl 2003, Wang, Blei 2012, Neal 2000 (alg 8).  A proposed split has REM parameters that are only slightly different than those for the original block.  Merges of two sets of REM parameters are performed by using the parameter for one blocks, and doing so for each dimension independently.  By keeping track of the probability of each of these transitions we can correctly compute the Metropolis-Hastings acceptance probability needed to guarantee we have an MCMC transition that leaves the posterior distribution invariant.

This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.  We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg. 8).

%$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$

Next pick two nodes $i,j$ uniformly.  If they belong to the same block $k$ (i.e. $z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
\begin{itemize}
%\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
\item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
\item Let $S$ be all nodes assigned to $k$.  Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
\end{itemize}

  If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e. $z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
\begin{itemize}
\item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$
\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
\item Let $S$ be the set of nodes assigned to $k$ or $l$. Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
\end{itemize}

The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})}\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}\frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$

The first two terms represent the prior. For the proposed merge step, note $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge}) = .5^{P(2K-3)}$ where $K$ is the number of blocks before the move.  We can compute $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.  For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.  $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.  We have avoided the dimensionality of the model during the MH move by creating an empty cluster when merging and filling a (newly created) empty cluster when splitting.  One can return to the previous state with positive probability...
